---
title: "Assignment 4: Data Wrangling (Spring 2025)"
author: "Angelica Rodriguez"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.
6. Ensure that code in code chunks does not extend off the page in the PDF.


## Set up your session 

1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.

1b. Check your working directory.

1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Add the appropriate code to reveal the dimensions of the four datasets. 

```{r setup, message = FALSE}
#1a 
#install.packages(tidyverse)
library(tidyverse)
#install.packages("lubridate")
library(lubridate)
#install.packages("here")
library(here)

#1b 
getwd()
here::here()

#1c 
#Read the directory 
dir.exists("/home/guest/EDA_Spring2025/Data/Raw")
#read all the files available
list.files("/home/guest/EDA_Spring2025/Data/Raw")  

#Read the files CVS
epa_data1<- read.csv("/home/guest/EDA_Spring2025/Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors = TRUE)
epa_data2<- read.csv("/home/guest/EDA_Spring2025/Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE)
epa_data3.PM25 <- read.csv("/home/guest/EDA_Spring2025/Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors = TRUE)
epa_data4.PM25 <- read.csv("/home/guest/EDA_Spring2025/Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors = TRUE)

#2 
#To see the name of the columns
lapply(list(epa_data1, epa_data2, epa_data3.PM25, epa_data4.PM25), colnames)
#To see the charcaterisctics of the data ,min, median  and mean of the data 
lapply(list(epa_data1, epa_data2, epa_data3.PM25, epa_data4.PM25), summary)
# To see the characteristics of de data if they are a factor or a number 
lapply(list(epa_data1, epa_data2, epa_data3.PM25, epa_data4.PM25), str)
#To see the dimensions of the data numbers of columns and number of data available
lapply(list(epa_data1, epa_data2, epa_data3.PM25, epa_data4.PM25), dim)


```

>All four datasets should have the same number of columns but unique record counts (rows). Do your datasets follow this pattern?


## Wrangle individual datasets to create processed files.

3. Change the Date columns to be date objects.

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r questions 3 thru 4}
#3
#Change format as date Data set 1, 2, 3, and 4 

epa_data1$Date <- mdy(epa_data1$Date)
epa_data2$Date <- mdy(epa_data2$Date) 
epa_data3.PM25$Date <- mdy(epa_data3.PM25$Date) 
epa_data4.PM25$Date <- mdy(epa_data4.PM25$Date)


#review changes 
str(epa_data1$Date)
str(epa_data2$Date) 
str(epa_data3.PM25$Date)
str(epa_data4.PM25$Date)

#4

#Create a new Data visualizations 1, 2, 3, and 4  with the selected 
#columns: "Date", "DAILY_AQI_VALUE","Site.Name",  "AQS_PARAMETER_DESC", 
#"COUNTY", "SITE_LATITUDE",
#"SITE_LONGITUDE")

#A epa_data1.Modified

epa_data1.Modified <- select(epa_data1,"Date", 
  "DAILY_AQI_VALUE", "Site.Name", 
  "AQS_PARAMETER_DESC",
  "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")

#B epa_data2.Modified

epa_data2.Modified <- select(epa_data2,"Date", 
  "DAILY_AQI_VALUE", "Site.Name", 
  "AQS_PARAMETER_DESC",
  "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")

#C epa_data3.Modified

epa_data3.PM25.Modified <- select(epa_data3.PM25,"Date", 
  "DAILY_AQI_VALUE", "Site.Name", 
  "AQS_PARAMETER_DESC",
  "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")

#D epa_dat4.Modified

epa_data4.PM25.Modified <- select(epa_data4.PM25,"Date", 
  "DAILY_AQI_VALUE", "Site.Name", 
  "AQS_PARAMETER_DESC",
  "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")

#5

#To fill all cells in AQS_PARAMETER_DESC with "PM2.5" 
# in epa_data3.Modified = "EPAair_O3_NC2019_raw.csv" Modified                 

epa_data3.PM25.Modified <-epa_data3.PM25.Modified %>%
  mutate(AQS_PARAMETER_DESC="PM2.5")

#and epa_data4.PM25.Modified = "EPAair_PM25_NC2018_raw.csv" 

epa_data4.PM25.Modified <-epa_data4.PM25.Modified %>% 
  mutate(AQS_PARAMETER_DESC="PM2.5")


#6 Saving all four processed datasets

write_csv(epa_data1.Modified, file="/home/guest/EDA_Spring2025/Data/Processed/epa_data1.Modified")

write_csv(epa_data2.Modified, file="/home/guest/EDA_Spring2025/Data/Processed/epa_data2.Modified")

write_csv(epa_data3.PM25.Modified, file="/home/guest/EDA_Spring2025/Data/Processed/epa_data3.PM25.Modified")

write_csv(epa_data4.PM25.Modified, file="/home/guest/EDA_Spring2025/Data/Processed/epa_data4.PM25.Modified")


```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include only sites that the four data frames have in common:  
    
  "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue",  
  "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", 
  "West Johnston Co.", "Garinger High School", "Castle Hayne", 
  "Pitt Agri. Center", "Bryson City", "Millbrook School" 
  
  (the function `intersect` can figure out common factor levels - but it will include sites with missing site information, which you don't want...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r questions 7 thru 11}
#7 
#To verify that all columns are equal 
lapply(list(epa_data1.Modified, epa_data2.Modified, epa_data3.PM25.Modified,epa_data4.PM25.Modified), colnames)

#To combine all data sets 
EPAair_O3_PM25_NC1819_Processed<- rbind (epa_data1.Modified, epa_data2.Modified, epa_data3.PM25.Modified,epa_data4.PM25.Modified)

#8 

common_sites <- c(
  "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue",  
  "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", 
  "West Johnston Co.", "Garinger High School", "Castle Hayne", 
  "Pitt Agri. Center", "Bryson City", "Millbrook School")

# Create a new data set named "DataAirNC.Modified" use ´filter´ to combined site 
# name and common sites group the variables, summarize by mean 
# and mutate 
DataAirNC.Modified <-
EPAair_O3_PM25_NC1819_Processed %>%
filter(Site.Name %in% common_sites & !is.na(Site.Name)) %>%
group_by(Date, Site.Name, COUNTY, AQS_PARAMETER_DESC) %>%
summarize(mean.AQI = mean(DAILY_AQI_VALUE),
mean.latitude = mean(SITE_LATITUDE),
mean.longitude = mean(SITE_LONGITUDE)) %>%
mutate(Month = month(Date), Year = year(Date))

dim(DataAirNC.Modified)
colnames(DataAirNC.Modified)

#9 #Spread datasets such that AQI values for ozone and PM2.5 are in separate 
#columns. Each location on a specific date should now occupy only one row.

#USING SPREAD

DataAirNC.Spread <- spread(DataAirNC.Modified, AQS_PARAMETER_DESC, mean.AQI)

#With pivot 

DataAirNC.spread2PIVOT <- DataAirNC.Modified %>% 
  pivot_wider(names_from = AQS_PARAMETER_DESC, values_from = mean.AQI)

DataAirNC.spread2PIVOT <- pivot_wider(DataAirNC.Modified, 
        names_from = AQS_PARAMETER_DESC, values_from = mean.AQI)

#10 Call up the dimensions of your new tidy dataset.
dim(DataAirNC.spread2PIVOT)
#here()


#11
write_csv(DataAirNC.spread2PIVOT, file="/home/guest/EDA_Spring2025/Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")


```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r questions 12 and 13}

#12
# Create the summaries "summary_data_frame"

summary_data_frame <- 
  DataAirNC.spread2PIVOT %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(meanOzone = mean(Ozone),
    meanPM2.5 = mean(PM2.5))   %>%
  drop_na(meanOzone)


#13
dim(summary_data_frame)

#14 
#replacing `drop_na` with `na.omit`
summary_data_frame_NAOMIT <- 
  DataAirNC.spread2PIVOT %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(meanOzone = mean(Ozone),
    meanPM2.5 = mean(PM2.5))   %>%
  na.omit(meanOzone)

```

14. Why did we use the function `drop_na` rather than `na.omit`? Hint: replace `drop_na` with `na.omit` in part 12 and observe what happens with the dimensions of the summary date frame.

> Answer: "na.omit" cleans and delete all the rows related to N/A values that can delete relevant data if they are not dependent.   

